---
title: "Comparing enhancer datasets"
author: "Andrew Johnston"
date: "11/03/2017"
output:
  html_document:
  css: styles.css
toc: yes
toc_depth: 4
pdf_document:
  toc: yes
word_document: default
---

```{r packages, echo=TRUE, warning=FALSE, message = FALSE}
# load in libraries
library(data.table)
library(ggplot2)
library(ggthemes)
library(eulerr)

# set options
options(scipen=999, stringsAsFactors = FALSE)
setwd("/Volumes/home/greally-lab/Deepa_Andrew/Deepa-helptagging/CpG_Annotations/")
```


For CD4 T cells, there were a few options. 

1) Enhancer annotations based off Roadmap ChIP-seq data. 

2 ) The ATACseq peaks as called in:
Buenrostro, J. D., Giresi, P. G., Zaba, L. C., Chang, H. Y., & Greenleaf, W. J. (2013). Transposition of native chromatin for fast and sensitive epigenomic profiling of open chromatin, DNA-binding proteins and nucleosome position. Nature Methods, 10(12), 1213–1218. http://doi.org/10.1038/nmeth.2688


```{r load-enhancer-sets, echo=TRUE, warning=FALSE, message = FALSE}
# roadmap enhancers 
road_enh <- read.table("Final_Enhancer_hg38.bed")
road_enh <- as.data.frame(road_enh)
road_enh$length <- road_enh$V3 - road_enh$V2
road_enh$name <- "road"

# CD4 ATAC enh 
ATAC_enh <- read.table("CD4_ATAC_peaks_hg38_noProm.bed")
ATAC_enh$length <- ATAC_enh$V3 - ATAC_enh$V2
ATAC_enh$name <- "ATAC"

# Ensembl enh 
ens_enh <- read.table("ensembl_enhancer_hg38.bed")
ens_enh$length <- ens_enh$V3 - ens_enh$V2
ens_enh$name <- "ens"

# create a data frame of all of the lengths plus the dataset they came from
road_ATAC_ens_enh_length <- rbind(road_enh[,7:8], ATAC_enh[,7:8], ens_enh[,7:8])

# graph the distribution of the lengths
ggplot(data = road_ATAC_ens_enh_length, aes(y=length, x=name, col=name)) +
  geom_violin() +
  theme_tufte() +
  ggtitle("Distribution of annotation length")

# look at the summary statistics 
summary(road_enh$length)
summary(ATAC_enh$length)
sum((ATAC_enh$length<10000)) # 31703
length(ATAC_enh$length) # 31709, so only 4 above 10 Kb. Will remove these in the final enhancer set
summary(ens_enh$length)
```
Based on the summary of length, the ensembl enhancers look the most convincing. 

The following code creates an euler plot of the overlap of basepairs among the different enhancer sets. 
```{r euler-plot,echo=TRUE, warning=FALSE, message = FALSE}
# this is the scratch to find the different areas within the graph.
12051108-649154
5028581-649154
165231936-11401954-4379427- 649154

1011296-649154
19407555-11401954-362142 - 649154

23429470 - 4379427 - 362142 -649154

overlaps <- euler(c("Roadmap" = 148801401, "ATAC" = 6994305, "Ensembl" = 18038747,
               "Roadmap&ATAC" = 11401954, "Roadmap&Ensembl" = 4379427, 
               "ATAC&Ensembl" = 362142, "Roadmap&ATAC&Ensembl" = 649154),
             shape = "ellipse")
# how good is the fit of the shapes? lower values are better. 
overlaps$stress
overlaps$diagError

# plot the euler figure
plot(overlaps, main="Basepair overlap")
```

```{r CpG-load,echo=TRUE, warning=FALSE, message = FALSE}
# load in the pvalues for CpG methylation between the two groups
CpGs <- fread("../../SMITE_10_2017/11-2017-meth_for_smite_angleadj_mean_anglea.txt")
CpGs <- as.data.frame(CpGs)
dim(CpGs) # 1433614 4
head(CpGs)
hist(CpGs$p.value)

# the qvalues doesn't give a great distribution for determining differences among enhancer sets
library(qvalue)
hist(qvalue(CpGs$p.value)$qvalue)
```

Now to read in the intersection of the different enhancer sets with CpGs and then to look at the density of the pvalues within each annotation. Ideally, the distribution should be uniform between 0 and 1, but there's a peak around .6 as seen in the pvalues from the 
```{r CpG-intersections,echo=TRUE, warning=FALSE, message = FALSE}
# CpG_inter_road
CpG_inter_road <- fread("CpG_inter_Road.bed")
CpG_inter_road <- as.data.frame(CpG_inter_road)
head(CpG_inter_road)
CpGs_road_wPos <- merge(x= CpGs, y= CpG_inter_road, by.x="X.tid", by.y = "V4")
head(CpGs_road_wPos)
nrow(CpGs_road_wPos) # 196284
summary(CpGs_road_wPos$p.value)
hist(CpGs_road_wPos$p.value)
CpGs_road_wPos$Type <- "Roadmap"

# CpG_inter_ATAC
CpG_inter_ATAC <- fread("CpG_inter_ATAC.bed")
CpG_inter_ATAC <- as.data.frame(CpG_inter_ATAC)
head(CpG_inter_ATAC)
CpGs_ATAC_wPos <- merge(x= CpGs, y= CpG_inter_ATAC, by.x="X.tid", by.y = "V4")
head(CpGs_ATAC_wPos)
nrow(CpGs_ATAC_wPos) # 41923
summary(CpGs_ATAC_wPos$p.value)
hist(CpGs_ATAC_wPos$p.value)
CpGs_ATAC_wPos$Type <- "ATAC"

# CpG_inter_ens
CpG_inter_ens <- fread("CpG_inter_ens.bed")
CpG_inter_ens <- as.data.frame(CpG_inter_ens)
head(CpG_inter_ens)
CpGs_ens_wPos <- merge(x= CpGs, y= CpG_inter_ens, by.x="X.tid", by.y = "V4")
head(CpGs_ens_wPos)
nrow(CpGs_ens_wPos) # 16835
summary(CpGs_ens_wPos$p.value)
hist(CpGs_ens_wPos$p.value)
CpGs_ens_wPos$Type <- "Ensembl"

# CpG_inter_ens_TF
CpG_inter_ens_TF <- fread("CpG_inter_ens_TF.bed")
CpG_inter_ens_TF <- as.data.frame(CpG_inter_ens_TF)
head(CpG_inter_ens_TF)
CpGs_ens_TF_wPos <- merge(x= CpGs, y= CpG_inter_ens_TF, by.x="X.tid", by.y = "V4")
head(CpGs_ens_TF_wPos)
nrow(CpGs_ens_TF_wPos) # 16835
summary(CpGs_ens_TF_wPos$p.value)
hist(CpGs_ens_TF_wPos$p.value)
CpGs_ens_TF_wPos$Type <- "Ensembl_TF"

CpG_inter_anno <- rbind(CpGs_road_wPos[,c(4,10)], CpGs_ATAC_wPos[,c(4,10)],
                        CpGs_ens_wPos[,c(4,10)], CpGs_ens_TF_wPos[,c(4,10)])

# plot the denisty graph of the pvalues of CpGs found within each annotation set.
ggplot(data=CpG_inter_anno, aes(x=p.value, col=Type)) +
  geom_density() +
  theme_tufte() +
  ggtitle("Distribution of pvalues (from Fisher's exact)\n in the different enhancer annotations")
```
The ensembl annotation is more enriched in "diff" methylated CpGs.

Now I'll make the final list of enhancer annotations as the combination of the ATAC marks as well as the emsembl enhancer annotation. I won't use the ensembl TF binding annotation because there's no annotation as to what the TFs are. I also will filter out any annotations longer than 10kb. 
```{r make-final-anno,echo=TRUE, warning=FALSE, message = FALSE}
# filter out those higher than 10kb and remove extraneous columns
ATAC_enh_10kb <- ATAC_enh[ATAC_enh$length<=10000,1:6]
ens_enh_10kb <- ens_enh[ens_enh$length<=10000,1:6]

ATAC_ens_enh <- rbind(ATAC_enh_10kb,ens_enh_10kb)
write.table(ATAC_ens_enh,"ATAC_ens_enh_hg38_no10kb.bed", append = F, quote = F, 
            row.names = F, col.names = F, sep = "\t")

# read in the intersection of the final annotation and CpGs
ATAC_ens_CpG_intersect <- read.table("ATAC_ens_CpG_intersect.txt")
dim(ATAC_ens_CpG_intersect)
head(ATAC_ens_CpG_intersect)
ATAC_ens_CpG_intersect_2 <- ATAC_ens_CpG_intersect[,c(4,10)]

# remove the old ATAC and enhancer annotation
CpG_pos_2 <- CpG_pos[,-c(6:7)]
# merge the new annotation with the old annotation
CpG_pos_enh <- merge(x = CpG_pos_2, y = ATAC_ens_CpG_intersect_2, all.x = TRUE, by.x = "X.tid",
                     by.y = "V10")

colnames(CpG_pos_enh)[6] <- "Enhancer"
write.table(CpG_pos_enh, "CpGID_Prom_GB_Enh2_annot.txt", append = F, quote = F,
            col.names = T, row.names = F, sep = "\t", na = "NA")
```
